---
title: "Practical Time-Series Clustering for Messy Data in R"
author: "Jonathan Page"
date: "1/13/2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Abstract

Identifying patterns in a pool of event data is hard. Time-series clustering provides a bag of lenses to change the level of detail when viewing and analyzing a collection of event data. Jonathan Page will present time-series clustering techniques using the dtwclust package in R and share his experience applying these techniques to mobile banking data in Kenya.


Attendees will receive a handout with graphs of time-series to manually classify as an exercise. Following the classification exercise and the presentation of two important clustering techniques, Dynamic Time Warping with Sakoe-Chiba constraints and the more recent k-Shape clustering algorithm, he will display the assignments given by the respective algorithms. This exercise and the discussion about it gives the participants an intuition about the benefits and drawbacks of each algorithm.

The code for generating the presented time-series analysis will be posted to GitHub. Participants will also receive a handout with a step-by-step process taking them from messy event data, to organized time-series, to classified time-series, and finally to visualizations useful for presenting the resulting time-series clusters.

Keywords: time-series clustering, time-series visualization, classification, event data

## Data

```{r}
load("clean.RData")
```

## Visual summary

```{r}
library(tidyverse)
transactions %>% filter(amount > 0) %>%
  ggplot(aes(day_of_campaign, amount)) +
  geom_point(alpha = 0.1) +
  scale_y_log10(labels = scales::comma_format()) +
  scale_x_log10()
```


```{r}
qplot(transactions$amount) + scale_x_log10(labels = scales::comma_format())
```

```{r}
qplot(transactions$day_of_campaign) + scale_x_log10(labels = scales::comma_format())
```



```{r}
quantile(floor(transactions$day_of_campaign), probs = c(0.01, 0.025, 0.05, 0.95, 0.975, 0.99), na.rm = TRUE)
```

Looking at the distribution of transactions is one piece of this puzzle, but what is the distribution of the last day contributed to a campaign.

```{r}
last_days <- transactions %>%
  group_by(campaign_id) %>%
  summarize(last_day = max(day_of_campaign)) %>%
  mutate(day2 = last_day > 1)
last_days %>%
  ggplot(aes(last_day)) + stat_ecdf(geom = "step", pad = FALSE) + geom_vline(xintercept=30)
```

```{r}
qplot(last_days$last_day) + scale_x_log10()
```

```{r}
last_days %>% filter(!is.na(day2)) %>%
ggplot(aes(last_day)) + geom_histogram() + facet_wrap(~ day2) + scale_x_log10(labels = scales::comma_format())
```



```{r}
quantile(round(last_days$last_day), probs = seq(0.05, 0.95, 0.05), na.rm = TRUE)
```

Half of these campaigns are over within 6 days.

The value of transactions may also have a pattern over time that is important. What is the value
of staying open another day? We don't know anything about an individual's campaign costs. Each individual
campaign may have different investments in campaigning outside of the platform in addition to any
marketing on the platform.

Instead of the just last day of a campaign, what is the distribution of the day
where a given percentage of either campaign dollars or contribution transactions occur?

### Distribution of campaign progress

```{r}
balances <- transactions %>%
  group_by(campaign_id) %>%
  arrange(payment_date) %>%
  mutate(campaign_transaction = row_number(), campaign_balance = cumsum(amount))
```

Distribution of number of transactions in a campaign:
```{r}
balances %>%
  group_by(campaign_id) %>%
  summarize(transaction_count = n()) %>%
  select(transaction_count) %>%
  qplot() + scale_x_log10(labels = scales::comma_format())
```

There are a large number of campaigns with only one transaction. This is one class of campaign.

Number of campaigns with only one transactions:
```{r}
balances %>%
  group_by(campaign_id) %>%
  summarize(transaction_count = n()) %>%
  filter(transaction_count == 1) %>%
  summarise(`Campaigns with only One Transaction` = n())
```

Number of campaigns with a single contributor, but multiple transactions:
```{r}
balances %>%
  group_by(campaign_id, contributor) %>%
  summarize(transaction_count = n()) %>%
  group_by(campaign_id) %>%
  summarize(transaction_count = sum(transaction_count), contributors = n()) %>%
  filter(contributors == 1, transaction_count > 1) %>%
  summarise(`Campaigns with Single Contributor, but Multiple Transactions` = n())
```

```{r}
balances %>%
  group_by(campaign_id, contributor) %>%
  summarize(transaction_count = n()) %>%
  group_by(campaign_id) %>%
  summarize(transaction_count = sum(transaction_count), contributors = n()) %>%
  filter(contributors > 1) %>%
  summarise(`Campaigns with Multiple Contributors` = n())
```

```{r}
campaigns %>% 
  filter(days_old >= 30) %>% 
  select(campaign_id) %>%
  left_join(balances) %>%
  group_by(campaign_id, contributor) %>%
  summarize(transaction_count = n()) %>%
  group_by(campaign_id) %>%
  summarize(transaction_count = sum(transaction_count), contributors = n()) %>%
  filter(contributors > 1) %>%
  summarise(`Campaigns with Multiple Contributors, age >= 30 days` = n())
```



Total number of campaigns with transaction data:
```{r}
balances %>%
  distinct(campaign_id) %>%
  nrow()
```


## Creating Archetypes

An initial analysis of the data will make it clear if there are any obvious structural groups.
One such group is the campaigns with only one transaction. Another is any campaign with only
one contributor.

Total campaigns with transaction data: (3,491)

* One Transaction (1,431)
* One Contributor, More than one transaction (161)
* Campaigns with multiple contributors <30 days olds (140)
* Campaigns with multiple contributors >= 30 days old (1,759)

The last set, those campaigns with multiple contributors (and at least 30 days old) 
is the set we really need to analyze. The goal of this analysis is to produce archetypes 
grouping campaigns to surface some typical patterns in campaign contributions.

```{r}
multiple_trans <- campaigns %>% 
  filter(days_old >= 30) %>% 
  select(campaign_id) %>%
  left_join(balances) %>%
  group_by(campaign_id, contributor) %>%
  summarize(transaction_count = n()) %>%
  group_by(campaign_id) %>%
  summarize(transaction_count = sum(transaction_count), contributors = n()) %>%
  filter(contributors > 1) %>%
  select(campaign_id) %>%
  left_join(balances) %>%
  group_by(campaign_id) %>%
  mutate(
    ending_balance = sum(amount), 
    balance_progress = sapply(campaign_balance / ending_balance, FUN=function(x) min(max(x, 0), 1)),
    total_transactions = n(),
    transaction_progress = campaign_transaction / total_transactions
  )
```

At 30 days what is the distribution of transaction and balance progress?
```{r}
multiple_trans %>%
  filter(day_of_campaign < 30) %>%
  group_by(campaign_id) %>%
  arrange(campaign_transaction) %>%
  summarise(balance_progress = last(balance_progress), transaction_progress = last(transaction_progress)) %>%
  ggplot(aes(transaction_progress)) + geom_histogram()
```

By 30 days, 932 out of 1,759 are completed

```{r}
multiple_trans %>%
  filter(day_of_campaign < 31) %>%
  group_by(campaign_id) %>%
  arrange(campaign_transaction) %>%
  summarise(balance_progress = last(balance_progress), transaction_progress = last(transaction_progress)) %>%
  filter(transaction_progress == 1) %>%
  nrow()
```


## Time series

There are several time-series views into transactions overtime (collapsing everything here to daily frequency).

stock
- balance_progress (normalized)
- balance

flow
- amount
- amount/ending_balance
- campaign_transactions
- campaign_transactions/total_transactions

```{r}
multiple_trans %>%
  ggplot(aes(day_of_campaign, balance_progress, group = campaign_id)) +
  geom_line(alpha = 0.1) + xlim(0, 30)
```

```{r}
daily_series <- multiple_trans %>%
  mutate(day_of_campaign = floor(day_of_campaign) + 1) %>%
  filter(day_of_campaign <= 30) %>%
  group_by(campaign_id, day_of_campaign) %>%
  arrange(id) %>%
  summarise(
    balance = last(campaign_balance),
    balance_progress = last(balance_progress),
    amount = sum(amount),
    amount_prop = sapply(amount / first(ending_balance), FUN=function(x) min(max(x, 0), 1)),
    transactions = n(),
    transaction_progress = last(transaction_progress)
  ) %>% 
  filter(day_of_campaign > 0)
summary(daily_series)
```

### Balance

```{r}
library(ggthemes)
daily_series %>%
  ggplot(aes(day_of_campaign, balance, group = campaign_id)) +
  geom_line(alpha = 0.1) +
  scale_y_log10(labels = scales::comma_format()) +
  labs(x = "Day of campaign", y = element_blank(), title = "Campaign Balance (log-scale)") +
  theme_tufte()
```

### Balance Progress

```{r}
daily_series %>%
  ggplot(aes(day_of_campaign, balance_progress, group = campaign_id)) +
  geom_line(alpha = 0.1)
```

### Amount

```{r}
daily_series %>%
  ggplot(aes(day_of_campaign, amount, group = campaign_id)) +
  geom_line(alpha = 0.1) +
  scale_y_log10(labels = scales::comma_format())
```


### Amount per Ending Balance

```{r}
daily_series %>%
  ggplot(aes(day_of_campaign, amount_prop, group = campaign_id)) +
  geom_line(alpha = 0.1)
```

### Transactions

```{r}
daily_series %>%
  ggplot(aes(day_of_campaign, transactions, group = campaign_id)) +
  geom_line(alpha = 0.1) +
  scale_y_log10(labels = scales::comma_format())
```

### Transactions Progress

```{r}
daily_series %>%
  ggplot(aes(day_of_campaign, transaction_progress, group = campaign_id)) +
  geom_line(alpha = 0.1)
```

## Clustering

To perform clustering we need to reshape our data into a matrix of trajectories with each row representing
a campaign and each column representing a day of the campaign.

```{r}
balance_traj <- daily_series %>%
  filter(day_of_campaign %in% 1:30) %>%
  select(campaign_id, day_of_campaign, balance) %>%
  spread(day_of_campaign, balance) %>%
  mutate(`1` = coalesce(`1`, 0)) %>%
  remove_rownames() %>%
  column_to_rownames("campaign_id") %>%
  apply(1, FUN=zoo::na.locf) %>%
  t()
save(daily_series, balance_traj, file = "balance.RData")
```

```{r}
amount_traj <-  daily_series %>%
  filter(day_of_campaign %in% 1:30) %>%
  select(campaign_id, day_of_campaign, amount) %>%
  spread(day_of_campaign, amount, fill = 0) %>%
  remove_rownames() %>%
  column_to_rownames("campaign_id")
```



Figure of balance trajectories

```{r}
library(ggthemes)
balance_traj %>%
  as.data.frame() %>%
  rownames_to_column("campaign_id") %>%
  gather(day_of_campaign, balance, -campaign_id, convert = TRUE) %>%
  ggplot(aes(day_of_campaign, balance, group = campaign_id)) +
  geom_line(alpha = 0.1) +
  scale_x_continuous(breaks = 1:4 * 7) +
  scale_y_log10(labels = scales::comma_format()) +
  labs(x = "Day of campaign", y = element_blank(), title = "Campaign Balance (log-scale)") +
  theme_tufte()
```


```{r}
amount_traj %>%
  as.data.frame() %>%
  rownames_to_column("campaign_id") %>%
  gather(day_of_campaign, amount, -campaign_id, convert = TRUE) %>%
  ggplot(aes(day_of_campaign, amount, group = campaign_id)) +
  geom_line(alpha = 0.01) +
  scale_x_continuous(breaks = 1:4 * 7) +
  scale_y_log10(labels = scales::comma_format()) +
  labs(x = "Day of campaign", y = element_blank(), title = "Campaign Daily Amount (log-scale)") +
  theme_tufte()
```


### DTW

```{r}
library(dtwclust)

pc_dtw4 <- tsclust(balance_traj, k = 4L,
                  distance = "dtw_basic",
                  trace = TRUE, seed = 1234,
                  norm = "L2", window.size = 2L,
                  args = tsclust_args(cent = list(trace = TRUE)))
pc_dtw9 <- tsclust(balance_traj, k = 9L,
                  distance = "dtw_basic",
                  trace = TRUE, seed = 1234,
                  norm = "L2", window.size = 2L,
                  args = tsclust_args(cent = list(trace = TRUE)))
pc_dtw16 <- tsclust(balance_traj, k = 16L,
                  distance = "dtw_basic",
                  trace = TRUE, seed = 1234,
                  norm = "L2", window.size = 2L,
                  args = tsclust_args(cent = list(trace = TRUE)))

save(pc_dtw4, pc_dtw9, pc_dtw16, file = "pc_dtw.RData")
```

```{r}
plot(pc_dtw4, type = "c")
```

```{r}
pc_dtw4@centroids
pc_dtw4@cluster
pc_dtw4@datalist %>% as.data.frame() %>%
  gather(campaign_id, balance)
cluster_label <- function(cluster) {
  paste0(cluster, '\n', pc_dtw4@clusinfo[order(pc_dtw4@clusinfo$size, decreasing = TRUE), 1][as.numeric(cluster)], ' campaigns')
}

centroids <- pc_dtw4@centroids[order(pc_dtw4@clusinfo$size, decreasing = TRUE)] %>% 
  as.data.frame(col.names = 1:4) %>% 
  mutate(day_of_campaign = row_number()) %>%
  gather(cluster, balance, -day_of_campaign) %>% 
  mutate(cluster = parse_number(cluster))

balance_traj %>% 
  as.data.frame() %>% 
  rownames_to_column("campaign_id") %>%
  mutate(
    cluster = match(pc_dtw4@cluster, order(pc_dtw4@clusinfo$size, decreasing = TRUE)), 
    campaign_id = as.numeric(campaign_id)
  ) %>%
  gather(day_of_campaign, balance, -campaign_id, -cluster, convert = TRUE) %>%
  ggplot(aes(day_of_campaign, balance + 1, group=campaign_id)) + geom_line(alpha = 0.1) +
  geom_line(aes(day_of_campaign, balance + 1, group = NULL), data = centroids, color = "blue") +
  facet_wrap(~ cluster, labeller = labeller(cluster = cluster_label)) + scale_y_log10(labels = scales::comma_format()) +
  theme_tufte() + labs(x = "Day of campaign", y = element_blank(), title = "Campaign Balance (log scale)")

pc_dtw4@centroids[order(pc_dtw4@clusinfo$size, decreasing = TRUE)] %>% 
  as.data.frame(col.names = 1:4) %>% 
  mutate(day_of_campaign = row_number()) %>%
  gather(cluster, balance, -day_of_campaign) %>% 
  mutate(cluster = parse_number(cluster))

```



### Shape-based Distance

`k-Shape`

```{r}
pc_sbd4 <- tsclust(balance_traj, type = "p", k = 4L, seed = 1234,
                  distance = "sbd")
pc_sbd9 <- tsclust(balance_traj, type = "p", k = 9L, seed = 1234,
                  distance = "sbd")
pc_sbd16 <- tsclust(balance_traj, type = "p", k = 16L, seed = 1234,
                  distance = "sbd")
save(pc_sbd4, pc_sbd9, pc_sbd16, file = "pc_sbd.RData")
pc_sbd4
```

```{r}
plot(pc_sbd4, type="sc")
```

## Worksheet

Choose one series from each cluster (dtw, sbd).

```{r}
load("balance.RData")
load("pc_dtw.RData")
load("pc_sbd.RData")
library(tidyverse)
library(ggthemes)
library(grid)
library(gridExtra)

y_max <- max(balance_traj)
gtraj <- function(values) {
  qplot(1:30, values + 1, geom = "line") + theme_void() + scale_y_log10(limits = c(1, y_max)) +  theme(axis.line.x = element_line(color = '#CCCCCC'))
}

add_example <- function(X, candidates) {
  c(X, sample(setdiff(candidates, X), 1))
}

example_ids <- c() %>%
  add_example(rownames(balance_traj)[which(pc_dtw4@cluster == 1)]) %>%
  add_example(rownames(balance_traj)[which(pc_dtw4@cluster == 2)]) %>%
  add_example(rownames(balance_traj)[which(pc_dtw4@cluster == 3)]) %>%
  add_example(rownames(balance_traj)[which(pc_dtw4@cluster == 4)]) %>%
  add_example(rownames(balance_traj)[which(pc_sbd4@cluster == 1)]) %>%
  add_example(rownames(balance_traj)[which(pc_sbd4@cluster == 2)]) %>%
  add_example(rownames(balance_traj)[which(pc_sbd4@cluster == 3)]) %>%
  add_example(rownames(balance_traj)[which(pc_sbd4@cluster == 4)])

```

Answers for the handout
```{r}
library(cowplot)
answer_layout <-rbind(
  c(rep(1, 4), NA, rep(2, 4)),
  c(3:6, NA, 7:10),
  11:19,
  20:28,
  29:37,
  38:46,
  47:55,
  56:64,
  65:73,
  74,82
)

cluster_from_idx = function(X) {
  ((X - 11) %% 9) + 1
}

row_from_idx = function(X) {
  ((X - 11) %/% 9) + 1
}

dist_from_idx = function(X) {
  ifelse(
    cluster_from_idx(X) < 5, #dtw
    dtw_basic(
      pc_dtw4@centroids[[cluster_from_idx(X)]],
      balance_traj[example_ids[row_from_idx(X)],],
      window.size = 2L
    ),
    unlist(sbd(
      pc_sbd4@centroids[[cluster_from_idx(X) - 5]],
      balance_traj[example_ids[row_from_idx(X)],]
    ))
  )
}

# return TRUE if this is the smallest for the relevant group
dist_ids <- setdiff(11:82, 15 + 9*(0:7))
closest_from_idx = function(X) {
  ifelse(
    cluster_from_idx(X) < 5, #dtw
    cluster_from_idx(X) == pc_dtw4@cluster[which(example_ids[row_from_idx(X)] == names(pc_dtw4@datalist))],
    cluster_from_idx(X) - 5 == pc_sbd4@cluster[which(example_ids[row_from_idx(X)] == names(pc_sbd4@datalist))]
  )
}

gs <- vector("list", 82)
gs[[1]] <- textGrob("DTW")
gs[[2]] <- textGrob("SBD")
# header figures
gs[3:6] <- lapply(pc_dtw4@centroids, gtraj)
gs[7:10] <- lapply(pc_sbd4@centroids, gtraj)
# sampled trajectories
gs[15 + 9*(0:7)] <- apply(balance_traj[example_ids, ], 1, gtraj)
# dtw distances

gs[dist_ids] <- lapply(dist_ids, function(X) {
  if (closest_from_idx(X)) {
    return(textGrob(formatC(dist_from_idx(X), format = "e", digits = 2), gp=gpar(fill="grey")))
  } else {
    return(textGrob(formatC(dist_from_idx(X), format = "e", digits = 2), gp=gpar(col="grey")))
  }
})

pdf("handout_answers.pdf", width = 8.5, height = 11) 
plot_grid(plotlist = c(gs[3:6],textGrob, gs[7:82]), ncol = 9)
dev.off()
```

Handout answers
```{r}
library(cowplot)
answer_layout <-rbind(
  c(rep(1, 4), NA, rep(2, 4)),
  c(3:6, NA, 7:10),
  11:19,
  20:28,
  29:37,
  38:46,
  47:55,
  56:64,
  65:73,
  74,82
)

cluster_from_idx = function(X) {
  ((X - 11) %% 9) + 1
}

row_from_idx = function(X) {
  ((X - 11) %/% 9) + 1
}

dist_from_idx = function(X) {
  ifelse(
    cluster_from_idx(X) < 5, #dtw
    dtw_basic(
      pc_dtw4@centroids[[cluster_from_idx(X)]],
      balance_traj[example_ids[row_from_idx(X)],],
      window.size = 2L
    ),
    unlist(sbd(
      pc_sbd4@centroids[[cluster_from_idx(X) - 5]],
      balance_traj[example_ids[row_from_idx(X)],]
    ))
  )
}

# return TRUE if this is the smallest for the relevant group
dist_ids <- setdiff(11:82, 15 + 9*(0:7))
closest_from_idx = function(X) {
  ifelse(
    cluster_from_idx(X) < 5, #dtw
    cluster_from_idx(X) == pc_dtw4@cluster[which(example_ids[row_from_idx(X)] == names(pc_dtw4@datalist))],
    cluster_from_idx(X) - 5 == pc_sbd4@cluster[which(example_ids[row_from_idx(X)] == names(pc_sbd4@datalist))]
  )
}

gs <- vector("list", 82)
gs[[1]] <- textGrob("DTW")
gs[[2]] <- textGrob("SBD")
# header figures
gs[3:6] <- lapply(pc_dtw4@centroids, gtraj)
gs[7:10] <- lapply(pc_sbd4@centroids, gtraj)
# sampled trajectories
gs[15 + 9*(0:7)] <- apply(balance_traj[example_ids, ], 1, gtraj)
# dtw distances

gs[dist_ids] <- lapply(dist_ids, function(X) {
    return(rectGrob(width = unit(0.125, "in"), height = unit(0.125, "in")))
})
```

Handout without answers
```{r}
pdf("handout.pdf", paper = "letter", width = 0, height = 0) 
handout_content <- plot_grid(plotlist = c(gs[3:6],textGrob, gs[7:82]), ncol = 9)
title <- ggdraw() + draw_label("Practical Time-Series Clustering Handout (2018) - Jonathan Page", fontface='bold')
plot_grid(title, handout_content, ncol=1, rel_heights=c(0.1, 1)) # rel_heights values control title margins
dev.off()
```

