---
title: | 
  | Practical Time-Series Clustering
  | for Messy Data in R
author: |
  | Jonathan Page
  | University of Hawaii Economic Research Organization
  | uhero.hawaii.edu
date: "`r format(Sys.time(), '%d %B, %Y')`"
fontsize: 10pt
header-includes:
  - \usepackage{amsmath}
output:
  beamer_presentation:
    md_extensions: +simple_tables
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
```

## Problem Definition

GoFundMe-like company in Kenya needs to understand the typical activity
patterns in campaign contributions.

Creating a list of campaign archetypes will
allow them to reason about the effects of 
changes to their platform and campaign-specific
interventions.


## Outline

1. Messy data -> Matrix of trajectories
2. Dynamic Time Warping (DTW) clusters
3. k-Shape (or Shape-based) clusters

# Messy Data

## Data Cleaning Plan

1. Select tables and columns necessary for analysis
2. Remove test campaigns and users
3. Produce long table of features
4. Create trajectory matrix for each feature

## Raw Data  {.smaller}

Transactions

| campaign_id | contributor_id | amount | payment_time        |
| ----------- | -------------- | ------ | ------------------- |
| 1           | 1              | 64     | 2016-01-01 00:00:01 |
| 1           | 2              | 128    | 2016-01-01 06:00:02 |
| 1           | 3              | 256    | 2016-01-02 12:34:56 |
| 1           | 3              | 512    | 2016-01-03 06:54:32 |
| 1           | 2              | 1024   | 2016-01-05 07:53:10 |
| 2           | 2              | 2048   | 2016-01-07 23:59:59 |

Campaigns

| campaign_id | start_time          |
| ----------- | ------------------- |
| 1           | 2016-01-01 00:00:00 |
| 2           | 2016-01-05 12:00:00 |

## Intermediate Data  {.smaller}

Combined and aggregated (in-time)

| campaign_id | balance | amount | contributors | day_of_campaign |
| ----------- | ------- | ------ | ------------ | --------------- |
| 1           | 192     | 192    | 2            | 1               |
| 1           | 448     | 256    | 1            | 2               |
| 1           | 960     | 512    | 1            | 3               |
| 1           | 1984    | 1024   | 1            | 5               |
| 2           | 2048    | 2048   | 1            | 3               |


## Matrix of trajectories {.smaller}

Balance trajectories

\[ b_{it} \]

--- --- ----- ----- -----
192 448 960   960   1984 
0   0   2048  2048  2048 
--- --- ----- ----- -----

Amount trajectories

\[ a_{it} \]

--- --- ----- ----- -----
192 256 512   0     1024
0   0   2048  0     0   
--- --- ----- ----- -----

## Data Cleaning

```{r clean, eval=FALSE, echo=TRUE}
library(tidyverse)
transactions <- read_csv("transactions.csv")
end_of_time <- max(transactions$payment_date)
campaigns <- read_csv(
    "campaigns.csv", 
    na = c("", "0000-00-00 00:00:00")
  ) %>%
  mutate(
    days_old = as.numeric(difftime(
      end_of_time, 
      date_created, 
      units = "days"
    ))
  )
```


## Data Cleaning

```{r clean2, eval=FALSE, echo=TRUE}
transactions <- transactions %>%
  inner_join(
    campaigns %>% 
      select(campaign_id, date_created)
  ) %>%
  mutate(
    day_of_campaign = as.numeric(difftime(
      payment_date, 
      date_created, 
      units = "days"
    ))
  ) %>%
  select(-date_created)
```


## Data Cleaning

```{r clean3, eval=FALSE, echo=TRUE}
clean_transactions <- campaigns %>% 
  filter(days_old >= 30) %>% 
  select(campaign_id) %>% 
  left_join(transactions) %>%
  group_by(campaign_id) %>% 
  summarize(
    transaction_count = n(), 
    contributors = length(unique(contributor))
  ) %>%
  filter(contributors > 1) %>% 
  select(campaign_id) %>%
  left_join(transactions) %>%
  group_by(campaign_id) %>%
  arrange(payment_date) %>%
  mutate(campaign_balance = cumsum(amount))
```

## Daily Series

```{r clean4, eval=FALSE, echo=TRUE}
daily_series <- clean_transactions %>%
  mutate(day_of_campaign = floor(day_of_campaign) + 1) %>%
  filter(day_of_campaign <= 30) %>%
  group_by(campaign_id, day_of_campaign) %>%
  arrange(id) %>%
  summarise(
    balance = last(campaign_balance),
    amount = sum(amount),
    transactions = n(),
  ) %>% 
  filter(day_of_campaign > 0)
```


## Long-form Balance Trajectories

```{r long_balances, echo=FALSE, message=FALSE, warning=FALSE}
load("balance.RData")
library(tidyverse)
library(ggthemes)
daily_series %>%
  ggplot(aes(day_of_campaign, balance, group = campaign_id)) +
  geom_line(alpha = 0.1) +
  scale_y_log10(labels = scales::comma_format()) +
  labs(x = "Day of campaign", y = element_blank(), title = "Campaign Balance (log-scale)") +
  theme_tufte()
```

## From sparse series to full matrix

```{r balance_traj, eval=FALSE, echo=TRUE}
balance_traj <- daily_series %>%
  filter(day_of_campaign %in% 1:30) %>%
  select(campaign_id, day_of_campaign, balance) %>%
  spread(day_of_campaign, balance) %>%
  mutate(`1` = coalesce(`1`, 0)) %>%
  remove_rownames() %>%
  column_to_rownames("campaign_id") %>%
  apply(1, FUN=zoo::na.locf) %>%
  t()
```

## From sparse series to full matrix

```{r amount_traj, eval=FALSE, echo=TRUE}
amount_traj <-  daily_series %>%
  filter(day_of_campaign %in% 1:30) %>%
  select(campaign_id, day_of_campaign, amount) %>%
  spread(day_of_campaign, amount, fill = 0) %>%
  remove_rownames() %>%
  column_to_rownames("campaign_id")
```


## Balance Trajectories

```{r balance_trajectories, echo=FALSE, message=FALSE, warning=FALSE}
balance_traj %>%
  as.data.frame() %>%
  rownames_to_column("campaign_id") %>%
  gather(day_of_campaign, balance, -campaign_id, convert = TRUE) %>%
  ggplot(aes(day_of_campaign, balance, group = campaign_id)) +
  geom_line(alpha = 0.1) +
  scale_x_continuous(breaks = 1:4 * 7) +
  scale_y_log10(labels = scales::comma_format()) +
  labs(x = "Day of campaign", y = element_blank(), title = "Campaign Balance (log-scale)") +
  theme_tufte()
```

# Dynamic Time Warping

## Dynamic Time Warping

```{r dtw-example, echo=FALSE, message=FALSE, warning=FALSE}
library(dtwclust)
plot(dtw(balance_traj[25L,], balance_traj[26L,], keep.internals = TRUE, window.type="sakoechiba", window.size = 2), type = "two",
     offset = -1000, 
     match.col = "blue",
     xlab = "Day of Campaign", ylab = "Balance")
```


## Sakoe-Chiba Window

Local Cost Matrix

```{r, fig.asp=1}
dtwWindow.plot(sakoeChibaWindow, window.size = 2L, reference = 30, query = 30)
```


## DTW Code

```{r, eval=FALSE, echo=TRUE}
pc_dtw <- tsclust(balance_traj, 
                  k = 4L,
                  distance = "dtw_basic",
                  seed = 1234,
                  norm = "L2", 
                  window.size = 2L)
```

## Results, k=4

```{r dtw4, echo=FALSE, message=FALSE, warning=FALSE}
load("pc_dtw.RData")
plot_tsclust <- function(tsclust_model) {
  cluster_label <- function(cluster) {
    paste0(cluster, '\n', tsclust_model@clusinfo[order(tsclust_model@clusinfo$size, decreasing = TRUE), 1][as.numeric(cluster)], ' campaigns')
  }
  centroids <- tsclust_model@centroids[order(tsclust_model@clusinfo$size, decreasing = TRUE)] %>% 
    as.data.frame(col.names = 1:tsclust_model@k) %>% 
    mutate(day_of_campaign = row_number()) %>%
    gather(cluster, balance, -day_of_campaign) %>% 
    mutate(cluster = parse_number(cluster))
  balance_traj %>% 
    as.data.frame() %>% 
    rownames_to_column("campaign_id") %>%
    mutate(
      cluster = match(tsclust_model@cluster, order(tsclust_model@clusinfo$size, decreasing = TRUE)), 
      campaign_id = as.numeric(campaign_id)
    ) %>%
    gather(day_of_campaign, balance, -campaign_id, -cluster, convert = TRUE) %>%
    ggplot(aes(day_of_campaign, balance + 1, group=campaign_id)) + geom_line(alpha = 0.1) +
    geom_line(aes(day_of_campaign, balance + 1, group = NULL), data = centroids, color = "blue") +
    facet_wrap(~ cluster, labeller = labeller(cluster = cluster_label)) + scale_y_log10(labels = scales::comma_format()) +
    theme_tufte(base_family = "sans") + 
    labs(
      x = "Day of campaign", 
      y = element_blank(), 
      title = "Campaign Balance",
      subtitle = paste0(
        tsclust_model@type, " clusters, ", 
        tsclust_model@distance, " distance, ", 
        tsclust_model@centroid, " centroids, k = ", tsclust_model@k
      )
    )
}
plot_tsclust(pc_dtw4)
```


## Results, k=9

```{r dtw9, echo=FALSE, message=FALSE, warning=FALSE}
plot_tsclust(pc_dtw9)
```

## Results, k=16

```{r dtw16, echo=FALSE, message=FALSE, warning=FALSE}
plot_tsclust(pc_dtw16)
```

# k-Shape Clustering

## Shape-based distance

Cross-correlation with shift

\[ SBD(\vec{x}, \vec{y}) = 1 - \max_s  \left( \frac{ \vec{x}_{(s)} \cdot \vec{y} }{\sqrt{\lVert\vec{x}\rVert^2  \lVert\vec{y}\rVert^2 }} \right) \]

\[ \vec{x}_{(s)} =
    \begin{cases} 
      (\overbrace{0, \ldots, 0}^{|s|}, x_1, x_2, \ldots, x_{m-s}), & s \geq 0 \\
      (x_{1-s}, \ldots, x_{m-1}, x_m, \underbrace{0, \ldots, 0}_{|s|}), & s < 0
    \end{cases}
\]

Paparrizos J, Gravano L (2015). "k-Shape: Efficient and Accurate Clustering of Time Series." In Proceedings of the 2015 ACM SIGMOD International Conference on Management of Data, {\em SIGMOD '15}, pp. 1855-1870. ACM, New York, NY, USA. ISBN 978-1-4503-2758-9. doi:10.1145/2723372.2737793.

## SBD Code

```{r, eval=FALSE, echo=TRUE}
pc_sbd4 <- tsclust(
  balance_traj, 
  type = "p", 
  k = 4L, 
  seed = 1234,
  distance = "sbd"
)
```


## Results for k=4

```{r sbd4, echo=FALSE, message=FALSE, warning=FALSE}
load("pc_sbd.RData")
plot_tsclust(pc_sbd4)
```

## Results for k=9

```{r sbd9, echo=FALSE, message=FALSE, warning=FALSE}
plot_tsclust(pc_sbd9)
```

## Results for k=16

```{r sbd16, echo=FALSE, message=FALSE, warning=FALSE}
plot_tsclust(pc_sbd16)
```


## Resources

R packages:

* `dtwclust`
* `tidyverse`

## Handout Answers

```{r, echo=FALSE, message=FALSE, warning=FALSE}
load("balance.RData")
load("pc_dtw.RData")
load("pc_sbd.RData")
library(tidyverse)
library(grid)
library(cowplot)

y_max <- max(balance_traj)
gtraj <- function(values) {
  qplot(1:30, values + 1, geom = "line") + theme_void() + scale_y_log10(limits = c(1, y_max)) +  theme(axis.line.x = element_line(color = '#CCCCCC'))
}

add_example <- function(X, candidates) {
  c(X, sample(setdiff(candidates, X), 1))
}

example_ids <- c() %>%
  add_example(rownames(balance_traj)[which(pc_dtw4@cluster == 1)]) %>%
  add_example(rownames(balance_traj)[which(pc_dtw4@cluster == 2)]) %>%
  add_example(rownames(balance_traj)[which(pc_dtw4@cluster == 3)]) %>%
  add_example(rownames(balance_traj)[which(pc_dtw4@cluster == 4)]) %>%
  add_example(rownames(balance_traj)[which(pc_sbd4@cluster == 1)]) %>%
  add_example(rownames(balance_traj)[which(pc_sbd4@cluster == 2)]) %>%
  add_example(rownames(balance_traj)[which(pc_sbd4@cluster == 3)]) %>%
  add_example(rownames(balance_traj)[which(pc_sbd4@cluster == 4)])

answer_layout <-rbind(
  c(rep(1, 4), NA, rep(2, 4)),
  c(3:6, NA, 7:10),
  11:19,
  20:28,
  29:37,
  38:46,
  47:55,
  56:64,
  65:73,
  74,82
)

cluster_from_idx = function(X) {
  ((X - 11) %% 9) + 1
}

row_from_idx = function(X) {
  ((X - 11) %/% 9) + 1
}

dist_from_idx = function(X) {
  ifelse(
    cluster_from_idx(X) < 5, #dtw
    dtw_basic(
      pc_dtw4@centroids[[cluster_from_idx(X)]],
      balance_traj[example_ids[row_from_idx(X)],],
      window.size = 2L
    ),
    unlist(sbd(
      pc_sbd4@centroids[[cluster_from_idx(X) - 5]],
      balance_traj[example_ids[row_from_idx(X)],]
    ))
  )
}

# return TRUE if this is the smallest for the relevant group
dist_ids <- setdiff(11:82, 15 + 9*(0:7))
closest_from_idx = function(X) {
  ifelse(
    cluster_from_idx(X) < 5, #dtw
    cluster_from_idx(X) == pc_dtw4@cluster[which(example_ids[row_from_idx(X)] == names(pc_dtw4@datalist))],
    cluster_from_idx(X) - 5 == pc_sbd4@cluster[which(example_ids[row_from_idx(X)] == names(pc_sbd4@datalist))]
  )
}

gs <- vector("list", 82)
gs[[1]] <- textGrob("DTW")
gs[[2]] <- textGrob("SBD")
# header figures
gs[3:6] <- lapply(pc_dtw4@centroids, gtraj)
gs[7:10] <- lapply(pc_sbd4@centroids, gtraj)
# sampled trajectories
gs[15 + 9*(0:7)] <- apply(balance_traj[example_ids, ], 1, gtraj)
# dtw distances

gs[dist_ids] <- lapply(dist_ids, function(X) {
  if (closest_from_idx(X)) {
    return(textGrob(formatC(dist_from_idx(X), format = "e", digits = 2), gp=gpar(fill="grey")))
  } else {
    return(textGrob(formatC(dist_from_idx(X), format = "e", digits = 2), gp=gpar(col="grey")))
  }
})

plot_grid(plotlist = c(gs[3:6],textGrob, gs[7:82]), ncol = 9)
```
